---
title: "Covid Data Example"
author: "D. ODay"
date: "7/19/2021"
output: html_document
---
Make sur eto import tidyverse
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I will start by reading in the data from the four main csv files.

```{r get_jhu_data}
## Get current Data in the four files
#They all begin in the same way
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

filenames <- c("time_series_covid19_confirmed_global.csv", 
               "time_series_covid19_deaths_global.csv",
               "time_series_covid19_confirmed_US.csv", 
               "time_series_covid19_deaths_US.csv"
               )
urls <- str_c(url_in, filenames)
```

```{r import_data, message=FALSE}
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
us_cases <- read_csv(urls[3])
us_deaths <- read_csv(urls[4])
````
After looking at global_cases and global_death, I would like to tidy those 
datasets and put each variable (date, cases, deaths) in their own column.
Also, I don't need Lat and Long for the analysis I'm planning, so I will get
rid of those and rename Region and State to be more R friendly

4:02 of video